Type-Safe Modular Parsing

### 2017-01-14 paper/sections/CaseStudy.tex
Fix refs, spell check
Down to 25 pages

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \multicolumn{3}{ c| }{\bfseries SLOC} & \multicolumn{3}{ c| }{\bfseries Time (ms)} \\ \cline{2-7}
        \multicolumn{1}{|c|}{} & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% \\
      \hline
      \input{resources/CaseStudyTable.tex}
      \hline
    \end{tabular}
    \caption{Comparison of SLOC and execution time}
    \label{tab:comparison}
\end{table}

\paragraph{Comparison Results}
Table \ref{tab:comparison} shows results of the comparison.
The overall result is that 69.2\% of code is reduced using our
library, and our implementation is 42.7\% slower.

@@

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \ilyaimpl{} & \multicolumn{2}{ c| }{\ourimpl{}} & \multicolumn{2}{ c| }{\ilyalongest{}} & \multicolumn{2}{ c| }{\ourclass{}} \\ \cline{2-8}
        \multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{\bfseries Time} & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% \\
      \hline
        Arith & 741 & 913 & +23.2 & 793 & +7.0 & 932 & +25.8 \\
        Untyped & 770 & 1018 & +32.2 & 821 & +6.6 & 1007 & +30.8 \\
      \hline
        \multicolumn{1}{|c|}{\dots} & \multicolumn{7}{c|}{\dots} \\
      \hline
        FullRecon & 1094 & 1645 & +50.4 & 1161 & +6.1 & 1652 & +51.0 \\
        FullPoly & 1398 & 2086 & +49.2 & 1511 & +8.1 & 2019 & +44.4 \\
        FullOmega & 1451 & 2352 & +62.1 & 1582 & +9.0 & 2308 & +59.1 \\
      \hline
        Total & 21746 & 31024 & +42.7 & 23260 & +7.0 & 30795 & +41.6 \\
      \hline
    \end{tabular}
    \caption{Execution time of four versions}
    \label{tab:ext-comparison}
\end{table}

Table \ref{tab:ext-comparison} shows part of the comparison results of these four versions,
full table is available online []. It suggests that the difference of running time between
using object algebra and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth mentioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.

>>>>

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \multicolumn{3}{ c| }{\bfseries SLOC} & \multicolumn{3}{ c| }{\bfseries Time (ms)} \\ \cline{2-7}
        \multicolumn{1}{|c|}{} & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% & \ilyaimpl{} & \ourimpl{} & \bfseries (+/-)\% \\
      \hline
      \input{resources/CaseStudyTable.tex}
      \hline
      \multicolumn{7}{c}{}
    \end{tabular}
    \caption{Comparison of SLOC and execution time.}
    \label{tab:comparison}
\end{table}

\paragraph{Comparison Results}
Table \ref{tab:comparison} shows results of the comparison.
The overall result is that 69.2\% of code is reduced using our
approach, and our implementation is 42.7\% slower.

@@

\begin{table}
    \centering
    \begin{tabular}{|l|r|r|r|r|r|r|r|}
      \hline
        \multirow{2}{*}{\bfseries Calculus Name} & \ilyaimpl{} & \multicolumn{2}{ c| }{\ourimpl{}} & \multicolumn{2}{ c| }{\ilyalongest{}} & \multicolumn{2}{ c| }{\ourclass{}} \\ \cline{2-8}
        \multicolumn{1}{|c|}{} & \multicolumn{1}{c|}{\bfseries Time} & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% & \bfseries Time & \bfseries (+/-)\% \\
      \hline
        Arith & 741 & 913 & +23.2 & 793 & +7.0 & 932 & +25.8 \\
        Untyped & 770 & 1018 & +32.2 & 821 & +6.6 & 1007 & +30.8 \\
      \hline
        \multicolumn{1}{|c|}{\dots} & \multicolumn{7}{c|}{\dots} \\
      \hline
        FullRecon & 1094 & 1645 & +50.4 & 1161 & +6.1 & 1652 & +51.0 \\
        FullPoly & 1398 & 2086 & +49.2 & 1511 & +8.1 & 2019 & +44.4 \\
        FullOmega & 1451 & 2352 & +62.1 & 1582 & +9.0 & 2308 & +59.1 \\
      \hline
        Total & 21746 & 31024 & +42.7 & 23260 & +7.0 & 30795 & +41.6 \\
      \hline
        \multicolumn{8}{c}{}
    \end{tabular}
    \caption{Execution time of four implementations.}
    \label{tab:ext-comparison}
\end{table}

Table \ref{tab:ext-comparison} shows part of the comparison results of these four versions,
full table is available online. It suggests that the difference of running time between
using object algebra and class is little, roughly 1\%.
The usage of longest match combinator slows the performance by 7\%. The main reason of slower
execution time may be the overall structure of the modular parsing approach, because we indeed have
more intermediate function calls and method overriding. However, it is worth mentioning that
because of the memoization technique of Packrat parsers, we are only constant times
slower, the algorithmic complexity is still the same.

### 2017-01-14 paper/sections/Discussion.tex
Down to 25 pages

\begin{figure}
\lstinputlisting[linerange=31-59]{../Scala/Parser/src/PaperCode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

@@

@@

It is natural to keep overriding existing parsers, whereas another potential use of this pattern is to reuse old versions of a parser. For instance,
we have a \lstinline{NewParser} that extends \lstinline{ExtParser}, with some more cases added, but for the lambda case we want to go back to the untyped one.
It can be achieved by instantiating an instance of \lstinline{BaseParser} to obtain the old \lstinline{pLam}. However, we must have \inlinecode{lazy} modifier for all the \inlinecode{pLam} starting from \inlinecode{BaseParser}.

\huang{i finally managed to make it work, with two overriding. i think this case may not be very common and the solution is kind of unly. not sure whether to keep this case or not}\haoyuan{I don't know if this works or makes sense.}

\lstinputlisting[linerange=94-102]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=NEWPARSER_UNTYPEDLAM

@@

\lstinputlisting[linerange=160-170]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty-printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}~\cite{ernst01FP} which is unfortunately not supported in Scala. Nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

>>>>

@@

\begin{figure}[ht]
\lstinputlisting[linerange=31-59]{../Scala/Parser/src/PaperCode/Sec4OA/Code4.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
\caption{Code for parsing a language with multi-sorted syntax for types and expressions.}
\label{fig:multi}
\end{figure}

@@

\begin{comment}
It is natural to keep overriding existing parsers, whereas another potential use of this pattern is to reuse old versions of a parser. For instance,
we have a \lstinline{NewParser} that extends \lstinline{ExtParser}, with some more cases added, but for the lambda case we want to go back to the untyped one.
It can be achieved by instantiating an instance of \lstinline{BaseParser} to obtain the old \lstinline{pLam}. However, we must have \inlinecode{lazy} modifier for all the \inlinecode{pLam} starting from \inlinecode{BaseParser}.

\lstinputlisting[linerange=94-102]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=NEWPARSER_UNTYPEDLAM
\end{comment}

@@

\begin{figure}[t]
\lstinputlisting[linerange=160-170]{../Scala/Parser/src/PaperCode/Sec5/Code1.scala}% APPLY:linerange=LANGUAGE_COMPONENTS_VARLAMEXPR
\caption{Reuse two language components \lstinline{VarExpr} and \lstinline{TypedLam}.}\label{fig:compose-components}
\end{figure}

Such composition shows perfect modularity. The only
drawback is that the glue code of composition appears to be boilerplate. As shown above, we are combining ASTs, parsers and pretty-printers of
\lstinline{VarExpr} and \lstinline{TypedLam} respectively. Such a pattern refers to \textit{family polymorphism}~\cite{ernst01FP} which is unfortunately not supported in Scala. Nonetheless, one can avoid such boilerplate by using metaprogramming techniques.

### 2017-01-14 paper/sections/ObjectAlgebras.tex
Down to 25 pages

\lstinputlisting[linerange=23-33]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use fields defined by keyword \inlinecode{def}, can not use those defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

\paragraph{Extensions}
Now let's try this pattern with the variables extension:
\lstinputlisting[linerange=37-44]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER

@@

\lstinputlisting[linerange=48-51]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

>>>>

\lstinputlisting[linerange=23-30]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=BASE_OA_PARSER
This is precisely the pattern that we advocate for modular parsing.
One important remark is we introduce \inlinecode{pE} for recursive calls.
The reason why we use an extra and seemingly redundant field, is due to a subtle issue caused by Scala language and its parser combinator library. There is a restriction of \inlinecode{super} keyword in Scala that \inlinecode{super} can only use fields defined by keyword \inlinecode{def}, can not use those defined by \inlinecode{val}, while the parser combinator library suggests using \inlinecode{val} to define parsers, especially for left-recursive ones. Our workaround is that we use different synonyms for \inlinecode{pE} in different traits, so that we can directly distinguish them by names without using \inlinecode{super}.

\paragraph{Extensions}
Now let's try this pattern with the variables extension:

\lstinputlisting[linerange=34-39]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER

@@

\lstinputlisting[linerange=43-46]{../Scala/Parser/src/PaperCode/Sec4OA/Code3.scala}% APPLY:linerange=EXT_OA_PARSER_CLIENT

### 2017-01-14 paper/sections/PackratParsing.tex
Down to 25 pages

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1-2-3$ is parsed as $(1-2)-3$ in the left-recursive
grammar, but after rewrite the result is $1((-2)-3))$\bruno{??This is
  not a valid expression}. The parse tree
must be transformed to recover the correct syntactic structure.

>>>>

Another issue of left-recursion elimination is that it requires extra
bookkeeping work to retain the original semantics. For example, the
expression $1-2-3$ is parsed as $(1-2)-3$ in the left-recursive
grammar, but after rewrite the information of left associativity is lost. The parse tree
must be transformed to recover the correct syntactic structure.
